---
title: "R-Ladies Rcpp Introduction"
author: |
  | Erin Rossiter
  | Washington University in St. Louis
  | erinrossiter@wustl.edu
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is Rcpp?

- R package
- Provides matching C++ classes for a large number of basic R data types
- Pass in R objects to C++ functions which return R objects!
- C++ is much, much faster than R
- Example: `catSurv`
    - some functions in packages written in R verses Rcpp/C++ are 10,000 times slower!
    - 10 vs .001 seconds... 10 seconds is too slow for real-time surveys

## Goals for tonight

1. Get Rcpp up and running on your computer (hopefully!!)
2. Walk through example code
    - Ways to set up an Rcpp function
    - Some C++ basics
3. Practice problems


## 1. Install Rcpp

Rcpp isn't installed as easily as other R packages.  You need to have compiler tools for R. (I'm not a computer scientist, so I don't fully understand the intricacies of this.)  So before we can use the Rcpp functions, we need to install a compiler.  This is different for Mac vs PC.  


### Mac

Steps:

1. Type `xcode-select --install` in the terminal **or** go to the App Store and install Xcode by hand
2. Verify installation by typing `gcc --version` in the terminal

<!--
I got help on these instructions from
http://thecoatlessprofessor.com/programming/r-compiler-tools-for-rcpp-on-os-x/
-->

### PC

Requirements:

1. Use Windows 7 or greater
2. Use R version 3.3.0 or greater
3. Be able to launch an application via “Run as Administrator” prompt


Steps:

1. Make sure R is installed on a path *without* spaces
2. Get the compatible version of Rtools for your version of R -- 3 ways to do this:
  - Way 1:
    - Open a C++ file in RStudio: File > New File > C++ File
    - Should prompt you to install Rtools.exe
    - With .exe in hand, make sure you following the prompts as shown in steps 4 - 14 from http://thecoatlessprofessor.com/programming/rcpp/install-rtools-for-rcpp/
  - Way 2:
    - Install and load the `installr` package
    - Run `install.Rtools()`
  - Way 3:
    - Download by hand here: https://cran.r-project.org/bin/windows/Rtools/
    - With .exe in hand, make sure you following the prompts as shown in steps 4 - 14 from http://thecoatlessprofessor.com/programming/rcpp/install-rtools-for-rcpp/
    

<!--
I got help on these instructions from
http://thecoatlessprofessor.com/programming/rcpp/install-rtools-for-rcpp/
-->

### Try it out

Run this chunk of code to install and load the Rcpp package, source in a few functions I wrote, and execute one.  If a (non-error) message prints, you're ready to go!

```{r warning=FALSE}
#install.packages("Rcpp")
library("Rcpp")

## Functions should appear in global environment
sourceCpp(file = "rcpp_source.cpp")

r_ladies()
```



## 2. Example Code

### A C++ Primer

> Please note, many of the following examples are not my own.  They're from Hadley Wickham's book, Advanced R, found online here: http://adv-r.had.co.nz/Rcpp.html.

Coding in C++ differs from R in several major ways:

1. Must specify data types for everything
2. Line breaks specified with a semicolon
3. For our purposes, there's no vectorized operations
4. C++ vector indices **start at zero**


```{r}
## Basic c++ function format
# return_type function_name(arg1_type arg1, arg2_type arg2, ...){
#   code;
#   code;
#   return somthing;
# }

cppFunction('int add1(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')

add1(1, 2, 3)
add1(1.2, 2, 3)
add1(TRUE, 2, 3)

## or for simple operations
evalCpp("1 + 2 + 3")
```

```{r}
## Equivalent R and C++ functions
## Returns  1 if the input is positive and -1 if it’s negative
signR <- function(x) {
  if (x > 0) {
    1
  } else if (x == 0) {
    0
  } else {
    -1
  }
}

cppFunction('int signC(int x) {
  if (x > 0) {
    return 1;
  } else if (x == 0) {
    return 0;
  } else {
    return -1;
  }
}')

signR(5)
signC(5)

evalCpp("5 > 0")
```


### Stand-Alone C++


There's another way to write and source in a C++ function: write all C++ source code in another file and use `sourceCpp(file = foo.cpp)` as we did above.

```{r}
add2(1, 2, 3) ## from file I provided, rcpp_source.cpp 
```

I prefer writing functions in .cpp files over writing inline C++ code with `cppFunction()`.  Writing stand-alone C++ code allows you to use the helpful text editor RStudio supplies (e.g., catches forgotten semicolons, auto indents, pretty colors, etc.)!  However, it's a bit more complicated.

Make sure:

1. Your file has a .cpp extension
2. The file starts with these 2 lines:

    `#include <Rcpp.h>`
    
    `using namespace Rcpp;`
    
3. Each function is preceeded by
    ` // [[Rcpp::export]]`
    
Let's check out rcpp_source.cpp.


### More Examples

Some Rcpp data types:

- `int`
- `double`
- `bool`
- `String`
- `NumericVector`
- `IntegerVector`
- `CharacterVector`
- `LogicalVector`
- `NumericMatrix`
- `IntegerMatrix`
- `CharacterMatrix`
- `LogicalMatrix` 
- `List`
- `DataFrame`
- ...and more

Let's check out rcpp_source.cpp for example functions using some of these types.

Note: you can also just use C++ data types! I prefer this as there's a ton of C++ documentation, help files, Q&A, and example code online.  There's good documentation for Rcpp as well (e.g., Hadley Wickham's *Advanced R* chapter), just not as much.  

```{r}
sumC(1:10)

check_greater(1:10, 5)

fill_matrix(2, 2)
fill_matrix(5, 5)

set.seed(19813)
x <- rnorm(1000, 10, 2)
y <- .25*x + rnorm(1000)
mod <- lm(y ~ x)
mod
grab_alpha(mod)

our_list <- list_of_vectors()
our_list$numeric
our_list$integer
our_list$logical
our_list$string
```



### Comparing Speed

```{r}
# install.packages("microbenchmark")
library(microbenchmark)

sumR <- function(x){
  total <- 0
  for(i in x){
    total <- total + i
  }
  return(total)
}

## Built-in sum() is unsurprisingly fast, and C++ looping is close!
## Comparable R function is much slower.
microbenchmark(sum(1:100),
               sumC(1:100),
               sumR(1:100))


## Even at something R is good at, C++ is faster (trivial example, but still!)
test_vec <- 1:10000
microbenchmark(sum(test_vec^5),
               pow_and_sum(test_vec, 5),
               unit = "s")
```


### Additional Functionality (if time)

#### Including R blocks in .cpp file.

Helpful for development/testing.  I've never done this.  Add the following to your .cpp file.  Like a code chunk in Rmarkdown.


```{r eval = FALSE, echo = TRUE}
/*** R
foo() ## whatever you would call from R!
*/
```


#### Calling R functions in a C++ function

Perhaps useful, but I think its tempting to be lazy and use this, potentially opening up a can of worms in your code!

```{r}
using_rfunc(sum, 1:10)
using_rfunc(diff, 1:10)
using_rfunc(cumsum, 1:10)
```




## 3. Practice 

#### Working with datatypes

Write a C++ function called `listC` that returns a list of length 3.

The list contains:

  - the vector [1,2,3]
  - a 2 by 2 matrix of zeros
  - a scalar 10
    
```{r}
## remember to always source in file with new function/edits!
sourceCpp(file = "rcpp_source.cpp")

## does it work?
# question1 <- listC();
```


#### Indexing and Matrices

Write a C++ function called `rowMeansC` to calculate column means of a matrix.  Assume no missing data. (Hint: if your matrix is `x`, `x.ncol()` and `x.nrow()` will give integer of number of columns and rows)

```{r}
## remember to always source in file with new function/edits!
sourceCpp(file = "rcpp_source.cpp")

set.seed(91308)
test_mat <- matrix(sample(1:100, replace = F), nrow = 10, ncol = 10)

## built-in R function
rowMeans(test_mat)

## does yours match?
# rowMeansC(test_mat)
```



