---
title: "R-Ladies Rcpp Introduction"
author: |
  | Erin Rossiter
  | Washington University in St. Louis
  | erinrossiter@wustl.edu
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is Rcpp?

- R package
- Provides matching C++ classes for a large number of basic R data types
- Pass in R objects to C++ functions which return R objects!
- C++ is much, much faster than R
- Example: `catSurv`
    - some functions in packages written in R verses Rcpp/C++ are 10,000 times slower!
    - 10 vs .001 seconds... 10 is too slow for real-time surveys

## Goals for tonight

1. Get Rcpp up and running on your computer (hopefully!!)
2. Walk through prepared, example code
3. Walk through how to set up your own code
4. Practice activity


## 1. Install Rcpp

Rcpp isn't installed as easily as other R packages.  You need to have compiler tools for R. (I'm not a computer scientists, so I don't fully understand the intricacies of this.)  So before we can use the Rcpp functions, we need to install a compiler.  This is different for Mac vs PC.  


### Mac

Steps:

1. Type `xcode-select --install` in the terminal **or** go to the App Store and install Xcode by hand
2. Verify installation by typing `gcc --version` in the terminal

<!--
I got help on these instructions from
http://thecoatlessprofessor.com/programming/r-compiler-tools-for-rcpp-on-os-x/
-->

### PC

Requirements:

1. Use Windows 7 or greater
2. Use R version 3.3.0 or greater
3. Be able to launch an application via “Run as Administrator” prompt


Steps:

1. Make sure R is installed on a path *without* spaces
2. Get the compatible version of Rtools for your version of R -- 3 ways to do this:
  - Way 1:
    - Open a C++ file in RStudio: File > New File > C++ File
    - Should prompt you to install Rtools.exe
    - With .exe in hand, make sure you following the prompts as shown in steps 4 - 14 from http://thecoatlessprofessor.com/programming/rcpp/install-rtools-for-rcpp/
  - Way 2:
    - Install and load the `installr` package
    - Run `install.Rtools()`
  - Way 3:
    - Download by hand here: https://cran.r-project.org/bin/windows/Rtools/
    - With .exe in hand, make sure you following the prompts as shown in steps 4 - 14 from http://thecoatlessprofessor.com/programming/rcpp/install-rtools-for-rcpp/
    

<!--
I got help on these instructions from
http://thecoatlessprofessor.com/programming/rcpp/install-rtools-for-rcpp/
-->

### Try it out

Run this chunk of code to install and load the Rcpp package, source in a few functions I wrote, and execute one.  If a (non-error) message prints, you're ready to go!

```{r warning=FALSE}
#install.packages("Rcpp")
library("Rcpp")

sourceCpp(file = "rcpp_source.cpp")

r_ladies()
```


## 2. Example Code

### A C++ Primer

Coding in C++ differs from R in 3 major ways:

1. Must specify data types for everything
2. Line breaks specified with a semicolon
3. For our purposes, there's no vectorized operations


```{r}
## Basic c++ function format
# return_type function_name(arg1_type arg1, arg2_type arg2, ...){
#   code;
#   code;
#   return somthing;
# }

cppFunction('int add1(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')

add1(1, 2, 3)
add1(1.2, 2, 3)
add1("hi", 2, 3)

add2(1, 2, 3)

evalCpp("1 + 2 + 3")
```

### Comparing Speed

```{r}
# install.packages("microbenchmark")
library(microbenchmark)

test_vec <- 1:10000
microbenchmark(sum(test_vec^5), pow_and_sum(test_vec, 5), unit = "s")
```

